First paper - Draft

### 2017-06-16 paper/paper.Rnw
Moved code into paper - everything good

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # For pisa_countrynames
  library(countrycode) # For region variable
  library(car)
  library(readr)
  library(SAScii)
  library(inequalityintsvy)
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)

@@

<<>>=
selected_cnt <- map(adapted_year_data, ~ select(.x, country, high_edu_broad, new_hisced, escs_dummy))

full_cnt <- 
  selected_cnt %>%
  enframe() %>%
  unnest(value) %>%
  mutate(year = as.character(dplyr::recode(.$name,
                `1` = 2000,
                `2` = 2003,
                `3` = 2006,
                `4` = 2009,
                `5` = 2012,
                `6` = 2015)),
         escs_dummy = as.character(escs_dummy))

sample_size <-
  full_cnt %>%
  count(year, country)

high_low_sample <-
  full_cnt %>%
  count(year, country, new_hisced) %>%
  filter(new_hisced %in% c(1, 7)) %>%
  spread(new_hisced, n) %>%
  rename(low_isced = `1`, high_isced = `7`)

ses_sample <-
  full_cnt %>%
  count(year, country, new_hisced, escs_dummy) %>%
  filter(new_hisced == 7 & escs_dummy != 0 | new_hisced == 1 & escs_dummy != 1) %>%
  ungroup() %>%
  select(-new_hisced) %>%
  spread(escs_dummy, n) %>%
  rename(low_low_isced = `0`, high__high_isced = `1`)

selected_cnts <-
  left_join(sample_size, high_low_sample) %>%
  left_join(ses_sample) %>%
  arrange(year, country)

@

<<eval = F>>=
# In this chunk you can join reading and math datasets
descrip_math <- map(results_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))

map2(results_math, results_read, inner_join, by = "country")
@

<<ci_for_difference>>=
reduced_data <-
    map2(results_math, years, function(.x, .y) {
      .x %>%
        mutate(wave = .y) %>%
        filter(!is.na(escs_dummy))
    }) %>%
    bind_rows() %>%
    mutate(lower = Mean - 1.96 * s.e.,
           upper = Mean + 1.96 * s.e.)
@

<<graphin differences, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
reduced_data %>%
    filter(country %in% c("United Kingdom")) %>%
    ggplot(aes(as.character(wave), Mean, group = escs_dummy, colour = escs_dummy)) +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
    geom_line() +
    facet_wrap(~ country)
@

<<>>=
# Calculate the joint standard error of the different
  se_data <-
    reduced_data %>%
    select(country, escs_dummy, wave, s.e.) %>%
    spread(escs_dummy, s.e.) %>%
    transmute(country, wave,
              se_diff = sqrt(abs(`1`^2 - `0`^2)))

@

<<include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

@@

countries <- c("Chile", "Austria", "Belgium", "Canada", "Czech Republic", "Denmark",
                 "Finland", "France", "Germany", "Italy", "Japan", "Netherlands", "Norway",
                 "Poland", "Spain", "Sweden", "United Kingdom", "United States")

reduced_data %>%
    select(country, escs_dummy, Mean, wave) %>%
    spread(key = escs_dummy, value = Mean) %>%
    mutate(diff = `1` - `0`) %>%
    gather(escs_dummy, Mean, -country, -wave, -`1`, -`0`) %>%
    left_join(se_data) %>%
    mutate(continent = countrycode(country, "country.name", "continent"),
           region = countrycode(country, "country.name", "region"),
           lower = Mean - 1.96 * se_diff,
           upper = Mean + 1.96 * se_diff) %>%
    filter(!is.na(continent), !is.na(region), country %in% vals) %>%
    ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
    geom_line() +
    geom_point(size = 0.5) +
    scale_colour_discrete(guide = F) +
    # geom_smooth(aes(group = region), colour = "blue", method = "lm") +
    coord_cartesian(ylim = c(0, 4)) +
    facet_wrap(~ country) +
    ggtitle("math")

@@

# No change:
# Canada - stable red - increase math
# UK - slight decrease red - stable math
# Belgium - no change
# Czech republic - no change
# Denmark no change
# Germany - no change
# ITaly - no change
# Japan -  no change
# Norway - no change
# Poland - no change
# Spain - no change

@

<<>>=
# Next steps:

# Show country patterns with reading and math together into one graph

# Making country groups of the graphs showing that some countries are increasing, others are steady
# While others are decreasing. Do this with a few countries, but end by showing all countries.

# Continue by studying what's happening in those countries where it's increasing/decreasing
# In the gap graph

@@

# The gap between the top 90th and the 50th, did it grow?
# The gap between the top 50th and the 10th, did it grow?

@@

# Show that within top and bottom educated categories, there is reasonable variation of SES index (to justify that your approach of suing the top and bottom percentiles within each category is the right approach)

# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?

@

>>>>

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(readr)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)

@@

<<merge_math_read>>=

# Function does a lot of things, but in short:

# Calculate the difference between the gap and together with it's joint s.e
# Also uncertainty intervals and returns a tibble with the difference between
# SES gaps with the adjusted SE difference + uncertainty intervals + the original
# data (the absolute numbers before the differences)

pisa_preparer <- function(df_math, df_read) {

descrip_math <- map(df_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
descrip_read <- map(df_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))

reduced_data_math <-
  map2(descrip_math, years, function(.x, .y) {
    .x %>%
      mutate(wave = .y) %>%
      filter(!is.na(escs_dummy))
  }) %>%
  bind_rows() %>%
  as_tibble() %>%
  mutate(lower_math = mean_math - 1.96 * se_math,
         upper_math = mean_math + 1.96 * se_math)

reduced_data_read <-
  map2(descrip_read, years, function(.x, .y) {
    .x %>%
      mutate(wave = .y) %>%
      filter(!is.na(escs_dummy))
  }) %>%
  bind_rows() %>%
  as_tibble() %>%
  mutate(lower_read = mean_read - 1.96 * se_read,
         upper_read = mean_read + 1.96 * se_read)

reduced_data <- left_join(reduced_data_math,
                          reduced_data_read, by = c("country", "escs_dummy", "wave"))

# Merging math and reading data
test_data <-
  reduced_data %>%
  select(country, wave, escs_dummy, contains("mean")) %>%
  gather(test, score, contains("mean"))

math_data <-
  reduced_data %>%
  select(country, wave, escs_dummy, contains("math")) %>%
  gather(test_bound, bound, contains("lower"), contains("upper")) %>%
  select(-contains("math")) %>%
  right_join(filter(test_data, test == "mean_math"))

read_data <-
  reduced_data %>%
  select(country, wave, escs_dummy, contains("read")) %>%
  gather(test_bound, bound, contains("lower"), contains("upper")) %>%
  select(-contains("read")) %>%
  right_join(filter(test_data, test == "mean_read"))

all_data <- bind_rows(math_data, read_data)

# Calculate the joint standard error of the difference
math_se_data <-
  reduced_data %>%
  select(country, escs_dummy, wave, se_math) %>%
  spread(escs_dummy, se_math) %>%
    transmute(country, wave,
              se_diff_math = sqrt(abs(`1`^2 - `0`^2)))

read_se_data <-
  reduced_data %>%
  select(country, escs_dummy, wave, se_read) %>%
  spread(escs_dummy, se_read) %>%
  transmute(country, wave,
            se_diff_read = sqrt(abs(`1`^2 - `0`^2)))

se_data <- left_join(math_se_data, read_se_data)

@@

math_diff <-
  reduced_data %>%
  select(wave, country, escs_dummy, mean_math) %>%
  spread(escs_dummy, mean_math) %>%
  transmute(wave, country, diff_math = `1` - `0`)

read_diff <-
  reduced_data %>%
  select(wave, country, escs_dummy, mean_read) %>%
  spread(escs_dummy, mean_read) %>%
  transmute(wave, country, diff_read = `1` - `0`)

data_summaries <-
  math_diff %>%
  left_join(read_diff) %>%
  left_join(se_data) %>%
  transmute(wave, country, diff_math, diff_read,
           lower_math = diff_math - 1.96 * se_diff_math,
           lower_read = diff_read - 1.96 * se_diff_read,
           upper_math = diff_math + 1.96 * se_diff_math,
           upper_read = diff_read + 1.96 * se_diff_read)

differences <-
  data_summaries %>%
  select(wave, country, diff_math, diff_read) %>%
  gather(test, difference, starts_with("diff")) %>%
  mutate(type_test = ifelse(.$test == "diff_math", "math", "read"))

bounds_lower <-
  data_summaries %>%
  select(wave, country, contains("lower")) %>%
  gather(lower_bound, lower, lower_math, lower_read) %>%
  mutate(type_test = ifelse(grepl("math", .$lower_bound), "math", "read"))

bounds_upper <-
  data_summaries %>%
  select(wave, country, contains("upper")) %>%
  gather(upper_bound, upper, upper_math, upper_read) %>%
  mutate(type_test = ifelse(grepl("math", .$upper_bound), "math", "read"))

# Getting the original data in
original_math <-
  reduced_data_math %>%
  select(wave, everything(), -se_math) %>%
  gather(metric, value, -(wave:escs_dummy)) %>%
  unite(combination, escs_dummy, metric, sep = "_") %>%
  spread(combination, value) %>%
  mutate(type_test = "math")

original_read <-
  reduced_data_read %>%
  select(wave, everything(), -se_read) %>%
  gather(metric, value, -(wave:escs_dummy)) %>%
  unite(combination, escs_dummy, metric, sep = "_") %>%
  spread(combination, value) %>%
  mutate(type_test = "read")

# final data
complete_data <-
  left_join(differences, bounds_lower) %>%
  left_join(bounds_upper) %>%
  left_join(original_math) %>%
  left_join(original_read)
}

complete_data_topbottom <- pisa_preparer(results_math, results_read)
complete_data_topmid <- pisa_preparer(results_math_topmid, results_read_topmid)
complete_data_midbottom <- pisa_preparer(results_math_midbottom, results_read_midbottom)

complete_data_topbottom <- mutate(complete_data_topbottom, type = "90th/10th SES gap")
complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES gap")
@

<<correlation_incomeineq, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
complete_data_topbottom %>%
  mutate(wave = as.character(wave)) %>%
  left_join(inequalityintsvy::economic_inequality, by = c("wave" = "year", "country")) %>%
  filter(indicators == "GINI") %>%
  group_by(country) %>%
  summarize(avg_diff = mean(difference, na.rm = T),
            avg_value = mean(value, na.rm = T)) %>%
  filter(avg_value < 8) %>%
  ggplot(aes(avg_value, avg_diff)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), linetype = "longdash", se = F)
@

<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

# 90/10 gaps acros countries
complete_data_topbottom %>%
  filter(country %in% c("United States", "Netherlands", "France",
                        "Germany", "Poland", "Finland")) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(-0.5, 3)) +
  facet_wrap(~ country)
@

<<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

# Comparing all gaps across countries
complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("United States", "Denmark", "France")) %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(-0.5, 3)) +
  facet_grid(country ~ type)
@

<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# Graphing how the top/bottom are evolving over time instead of absolute difference

complete_data_topbottom %>%
  bind_rows(complete_data_topmid) %>%
  bind_rows(complete_data_midbottom) %>%
  filter(country %in% c("Germany", "Denmark", "France"), type_test == "math") %>%
  select(wave, country, type_test, type, contains("math")) %>%
  mutate(type = factor(type,
                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
                       ordered = TRUE)) %>%
  gather(score, value, -(wave:type)) %>%
  separate(score, c("ses", "score"), sep = 2) %>%
  spread(score, value) %>%
  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "longdash") +
  geom_line() +
  geom_point(size = 0.5) +
  coord_cartesian(ylim = c(-0.5, 3)) +
  facet_grid(country ~ type)

@@

# No change:
# Canada - stable red - increase math
# UK - slight decrease red - stable math
# Belgium - no change
# Czech republic - no change
# Denmark no change
# Germany - no change
# ITaly - no change
# Japan -  no change
# Norway - no change
# Poland - no change
# Spain - no change
@

% How to output data frames are latex tables?
<<rate_change>>=

avg_increase_fun <- function(df, class) {

# Average standard deviation increase
  data_ready <-
    df %>%
    select(wave, country, type_test, contains("mean_math")) %>%
    gather(metric, value, -(wave:type_test)) %>%
    separate(metric, c("ses", "test"), sep = 2) %>%
    spread(test, value) %>%
    mutate(ses = gsub("_", "", ses)) %>%
    filter(type_test == "math", ses == class) %>%
    split(.$country) %>%
    map(~ mutate(.,
                 diff = c(diff(mean_math, lag = 1), NA),
                 perc = round(diff / mean_math, 2) * 100,
                 perc_pos = mean(perc > 0, na.rm = T))) %>%
    enframe() %>%
    unnest(value) %>%
    split(.$country)

    map2(data_ready, names(data_ready), ~ {

      print(.y)

      mean_df <-
        bootstrapper(.x, mean(diff, na.rm = T), B = 500) %>%
        filter(type == "bootstrap") %>%
        rename(mean = value)

      sd_df <-
        bootstrapper(.x, sd(diff, na.rm = T), B = 500) %>%
        filter(type == "bootstrap") %>%
        rename(sd = value)

      suppressMessages(
        left_join(mean_df, sd_df) %>%
        mutate(lower_bound = mean - 1 * sd,
               upper_bound = mean + 1 * sd)
      )
    }) %>%
    enframe() %>%
    unnest(value)
}

avg_sd_increase_high <- avg_increase_fun(complete_data_topbottom, 1)
avg_sd_increase_low <- avg_increase_fun(complete_data_topbottom, 0)
@

<<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

full_data <-
  left_join(avg_sd_increase_high, avg_sd_increase_low, by = "country") %>%
  mutate(continent = ifelse(country %in% countries, "my_cnt", "other_cnt"))

colnames(full_data) <- c("country", "high_increase", "low_increase", "continent")

lims <- list(xlim = c(-0.15, 0.25), ylim = c(-0.25, 0.25))

rect_data <- tibble(xst = c(lims$xlim[1], 0),
                    xen = c(0.0, lims$xlim[2]),
                    yst = c(0.0, lims$ylim[1]),
                    yen = c(lims$ylim[2], 0),
                    colour = c("red", "green"))

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1) +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES are catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES are increasing  \n faster than Low SES",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@

% How to output data frames are latex tables?
<<perc_increase_tables>>=

# Show the rates at which is increasing/decreasing
perc_increase_fun <- function(df) {

  # Average standard deviation increase
  data_ready <-
    df %>%
    select(wave, country, type_test, difference) %>%
    group_by(type_test) %>%
    split(.$country) %>%
    map(~ {
      .x <-
        spread(.x, wave, difference) %>%
        ungroup()

      year_vars <- sum(map_dbl(.x, is.numeric)) - 1
      years_subtract <- names(.x)[c(ncol(.x) - year_vars, ncol(.x))]
      years_subtract <- lapply(years_subtract, as.name)

      last_year <- rlang::new_quosure(years_subtract[[2]], env = .GlobalEnv)
      first_year <- rlang::new_quosure(years_subtract[[1]], env = .GlobalEnv)

      years_available <-
        .x %>%
        gather(year, val, -(country:type_test)) %>%
        group_by(type_test) %>%
        summarise(yr_avaible = sum(!is.na(val))) %>%
        pull(yr_avaible)

      year_sd <-
        .x %>%
        gather(year, val, -(country:type_test)) %>%
        split(.$type_test) %>%
        map_dbl(~ bootstrapper(.x, mad(val, na.rm = T), B = 100) %>% .[[2, 2]]) %>%
        round(2) * 100

      .x %>%
        transmute(type_test,
                  country,
                  perc_diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
                  sd_year = year_sd,
                  diff_lower = perc_diff - 1 * year_sd,
                  diff_upper = perc_diff + 1 * year_sd,
                  years_available = years_available)
    })
  data_ready
}

top_bottom_perc <- perc_increase_fun(complete_data_topbottom)
top_mid_perc <- perc_increase_fun(complete_data_topmid)
mid_bottom_perc <- perc_increase_fun(complete_data_midbottom)

# Gap is closing at an average of the variable diff per year.
@

<<next_steps>>=
# Next steps:

@@

@@

# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
@
